The Success of College Basketball Teams can be predicted using several advanced team statistics

### 2022-04-23 Paper-On-NCAAB-Stats/outputs/paper/Paper-On-NCAAB-Stats.Rmd
prelim model work

#Load in libraries
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
library(modelsummary)
```

@@

Even with these biases, the data should not be affected too substantially, and information can still be learned by analyzing it. This can be done using packages such as Tidyverse [@citeTidyverse], Dplyr [@citeDplyr], MASS [@citeMASS], and more in R [@citeR], where the data will be looked at graphically and modeled to answer the research question and tell a story about which team statistics play the biggest factor in team success. Before getting to the modeling, we ust first take a deeper look into the data that we will be using to do so.

@@

>>>>

#Load in libraries
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
library(modelsummary)
library(DiagrammeR)
library(tidymodels)
library(broom)
```

@@

Even with these biases, the data should not be affected too substantially, and information can still be learned by analyzing it. This can be done using packages such as Tidyverse [@citeTidyverse], Dplyr [@citeDplyr], modelsummary [@citeMS], and more in R [@citeR], where the data will be looked at graphically and modeled to answer the research question and tell a story about which team statistics play the biggest factor in team success. Before getting to the modeling, we ust first take a deeper look into the data that we will be using to do so.

@@

While we do have four strong candidate explanatory variables that may be used in the model, we still must decide what combination of these variables will make for the best possible model. First, it is a good idea to think logically if each of these variables should be included in the model, and the variable in particular we should analyze is offensive rating. Based on the visualizations analyzed in Section \@ref(data), it is clear that there is a relationship between offensive rating and a teams simple rating system, but that does not necessarily mean there is direct causality. Of course if a team scores more points and has a higher offensive rating, they will win more games and have a higher simple rating system value, but a high offensive rating is not what is causing a team to have success. It is instead underlying statistics that cause a team to score more points and thus have a higher simple rating system value. These underlying statistics, such as true shooting percentage, total rebounding percentage, and turnover percentage, affect both offensive rating and simple rating system and thus make it seem like offensive rating and SRS have a causal relationship. For example, teams with a higher true shooting percentages will hit more of their shots causing them to score more points, leading to a higher offensive rating, and thus causing them to win more games by larger amounts, and have a higher SRS value. So, these statistics are the true variables affecting simple rating system, and offensive rating is actually just acting as a mediator for the relationship between them. This relationship can be shown through the use of directed acrylic graphs (DAGs) created with DiagrammeR [@citeDiagramme], seen in Figure \@ref(fig:dags).

```{r dags, echo=FALSE, fig.cap="A DAG showing Offensive Rating as a mediator}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext, fontsize = 10, fontname = Helvetica]
    A [label = 'True Shooting % / Total Rebound % / Turnover %']
    B [label = 'Offensive Rating']
    C [label = 'Simple Rating System']
  edge [minlen = 2, arrowhead = vee]
    A -> C
    A -> B
    B->C
  { rank = same; A; C}
}
", height = 300)
```
In the DAG, it can be seen how true shooting percentage, total rebound percentage, and turnover percentage all affect and have causality with simple rating system. It can also be seen that offensive rating affects simple rating system too, but it is also affected by the three variables listed above and is being used as a mediator between these variables and simple rating system. As this mediator, offensive rating is explaining exactly why these three variables affect simple rating system [@citeBhandari]. The three variables affect team success because the three variables cause teams to either score more or less points on offense, giving them either a better or worse offensive rating, and thus causing them to either win or lose more games, which is essentially what the DAG is explaining. As a result of this, offensive rating should not be used in the model as the mediator will affect the strength of the relationship between the true predictor variables and simple rating system.

With the final three predictor variables, there are 7 possible combinations that can be used as predictors for a linear regression model. We can then test these models against one another to determine which model is the best at predicting simple rating system, and the variables used in the model will be the main team statistics that affect how much success a college basketball team has.

```{r}
library(tidymodels)

set.seed(9917)

ncaab_stats_split <- 
  initial_split(data = ncaab_stats, 
                prop = 0.80)

ncaab_stats_train <- training(ncaab_stats_split)

ncaab_stats_test <- testing(ncaab_stats_split)

```{r, include=FALSE}
mods <- list(
  "mod_ts_trb_tov" = lm(srs ~ ts_percent + trb_percent + tov_percent, data = ncaab_stats_train),
  "mod_ts_trb" = lm(srs ~ ts_percent + trb_percent, data = ncaab_stats_train),
  "mod_ts_tov" = lm(srs ~ ts_percent + tov_percent, data = ncaab_stats_train),
  "mod_trb_tov" = lm(srs ~ trb_percent + tov_percent, data = ncaab_stats_train),
  "mod_ts" = lm(srs ~ ts_percent, data = ncaab_stats_train),
  "mod_trb" = lm(srs ~ trb_percent, data = ncaab_stats_train),
  "mod_tov" = lm(srs ~ tov_percent, data = ncaab_stats_train)
)
```

```{r gof, echo=FALSE}
modelsummary(mods,
             output = "kableExtra",
             fmt = 2,
             title = "Comparing goodness of fit measures of several potential models predicting Simple Rating System. Column names represent which variables are included in each model.")
```

```{r}
fitted_and_resids <- 
  augment(mod_ts_trb_tov,
          data = ncaab_stats_train)

assump_plot1 <- ggplot(fitted_and_resids, aes(x = .resid)) + 
  geom_histogram(binwidth = 0.5, fill="black", colour="gray") +
  theme_classic() +
  labs(y = "Number of occurrences",
       x = "Residuals")

assump_plot2 <- ggplot(fitted_and_resids, aes(x = ts_percent, y = .resid)) + 
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
  theme_classic() +
  labs(y = "Residuals",
       x = "True Shooting %")

assump_plot3 <- ggplot(fitted_and_resids, aes(x = trb_percent, y = .resid)) + 
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
  theme_classic() +
  labs(y = "Residuals",
       x = "Total Rebound %")

assump_plot4 <- ggplot(fitted_and_resids, aes(x = tov_percent, y = .resid)) + 
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
  theme_classic() +
  labs(y = "Residuals",
       x = "Turnover %")

assump_plot5 <- ggplot(fitted_and_resids, aes(x = srs, .fitted)) + 
  geom_point() +
  theme_classic() +
  labs(y = "Estimated Simple Rating System",
       x = "Actual Simple Rating System")
```

```{r assumplots, fig.height=8, fig.cap="Residuals and Fitted Value plots for linear regression model predicting Simple Rating System using a training dataset on college basketball statistics}
ggarrange(assump_plot1,assump_plot5,assump_plot2,assump_plot3, assump_plot4, ncol=2, nrow=3)
```
